---
html_document:
keep_md: yes
output:
title: "Analysis of the Adverse Health and Economic Impacts of US Storms"
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

## Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

## Data processing
load required libraries
```{r}
library(dplyr) # load dplyr for data manipulation
library(ggplot2) # ggplot for data visualization
# install.packages("gridExtra")
library(gridExtra) # to plot graphs side by side
library(reshape2) # to melt the dataframe
```
Loading the data
```{r}
if(!file.exists("StormData.csv.bz2")) {
Original_Data_URL <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(Original_Data_URL, destfile="StormData.csv.bz2")
}
data <- read.csv("StormData.csv.bz2", stringsAsFactors=F)
head(data)
```
```{r}
# Subset (data) storm database
tidydata <- data[,c('EVTYPE','FATALITIES','INJURIES', 'PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP')]
# Convert H, K, M, B units to calculate Property Damage 
tidydata$PROPDMGNUM = 0
tidydata[tidydata$PROPDMGEXP == "H", ]$PROPDMGNUM = tidydata[tidydata$PROPDMGEXP == "H", ]$PROPDMG * 10^2
tidydata[tidydata$PROPDMGEXP == "K", ]$PROPDMGNUM = tidydata[tidydata$PROPDMGEXP == "K", ]$PROPDMG * 10^3
tidydata[tidydata$PROPDMGEXP == "M", ]$PROPDMGNUM = tidydata[tidydata$PROPDMGEXP == "M", ]$PROPDMG * 10^6
tidydata[tidydata$PROPDMGEXP == "B", ]$PROPDMGNUM = tidydata[tidydata$PROPDMGEXP == "B", ]$PROPDMG * 10^9
# Convert H, K, M, B units to calculate Crop Damage
tidydata$CROPDMGNUM = 0
tidydata[tidydata$CROPDMGEXP == "H", ]$CROPDMGNUM = tidydata[tidydata$CROPDMGEXP == "H", ]$CROPDMG * 10^2
tidydata[tidydata$CROPDMGEXP == "K", ]$CROPDMGNUM = tidydata[tidydata$CROPDMGEXP == "K", ]$CROPDMG * 10^3
tidydata[tidydata$CROPDMGEXP == "M", ]$CROPDMGNUM = tidydata[tidydata$CROPDMGEXP == "M", ]$CROPDMG * 10^6
tidydata[tidydata$CROPDMGEXP == "B", ]$CROPDMGNUM = tidydata[tidydata$CROPDMGEXP == "B", ]$CROPDMG * 10^9
```

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health ?

```{r}
# plot number of fatalities with the most harmful event type
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=tidydata, sum)
fatalities <- fatalities[order(-fatalities$FATALITIES), ][1:10, ]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)

ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Fatalities") + ggtitle("Number of fatalities by top 10 Weather Events")
```
```{r}
# plot number of injuries with the most harmful event type
injuries <- aggregate(INJURIES ~ EVTYPE, data=tidydata, sum)
injuries <- injuries[order(-injuries$INJURIES), ][1:10, ]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

ggplot(injuries, aes(x = EVTYPE, y = INJURIES)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Injuries") + ggtitle("Number of injuries by top 10 Weather Events")
```
2. Across the United States, which types of events have the greatest economic consequences?
```{r}
# plot number of damages with the most harmful event type
damages <- aggregate(PROPDMGNUM + CROPDMGNUM ~ EVTYPE, data=tidydata, sum)
names(damages) = c("EVTYPE", "TOTALDAMAGE")
damages <- damages[order(-damages$TOTALDAMAGE), ][1:10, ]
damages$EVTYPE <- factor(damages$EVTYPE, levels = damages$EVTYPE)

ggplot(damages, aes(x = EVTYPE, y = TOTALDAMAGE)) + 
    geom_bar(stat = "identity", fill = "blue", las = 3) + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
    xlab("Event Type") + ylab("Damages ($)") + ggtitle("Property & Crop Damages by top 10 Weather Events")

# The answer
"The greatest economic consequence event is" 
damages$EVTYPE[1]
```

